---
title: "Milestone Report for Capstone Project"
author: "Maria P. Frushicheva"
date: "August 12, 2016"
output: html_document
---
## Introduction 

__Importance:__ Using hospital quality ratings, patients are able to make a better decision in what hospital they want to be treated and where the best care is available in state of California, based on overall hospital performance or based on particular medical condition or procedure. 

__Question:__ Can we predict hospital quality ratings based on risk adjusted mortality rates, number of deaths, number of cases, medical procedures performed and medical conditions treated for 2012-2013?

## Description of Data Set

__Dataset:__ is available from [California Hospital Inpatient Mortality Rates and Quality Ratings, 2012-2013](https://chhs.data.ca.gov/Healthcare/California-Hospital-Inpatient-Mortality-Rates-and-/rpkf-ugbp). 

__Description of dataset:__ The dataset contains risk-adjusted mortality rates, and number of deaths and cases for 6 medical conditions treated (Acute Stroke, Acute Myocardial Infarction, Heart Failure, Gastrointestinal Hemorrhage, Hip Fracture and Pneumonia) and 6 procedures performed (Abdominal Aortic Aneurysm Repair, Carotid Endarterectomy, Craniotomy, Esophageal Resection, Pancreatic Resection, Percutaneous Coronary Intervention) in California hospitals for 2012 and 2013. This dataset does not include conditions treated or procedures performed in outpatient settings. 

## Description, Analysis and Cleaning of Variables in the Data Set

```{r,warning=FALSE,echo=FALSE,message=FALSE}
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("ggplot2")
#install.packages("gridExtra")
#install.packages("ggmap")
require(dplyr)
require(tidyr)
require(ggplot2)
require(gridExtra)
require(ggmap)
#library(maptools)
```

__Load the data from csv file.__ 
```{r,warning=FALSE,message=FALSE}
setwd("C:/Users/postdoc/Dropbox (Personal)/SpringBoard Fund/Rprojects/")
data <- read.csv("California_Hospital_Inpatient_Mortality_Rates_and_Quality_Ratings__2012-2013.csv",sep=",",header=TRUE)
df <- tbl_df(data)
# glimpse(df)
```

__Dataset:__ __`r nrow(df)`__ observations and __`r ncol(df)`__ variables. 

__Variables with missing values:__

- Risk Adjusted Mortality Rate: The Risk Adjusted Mortality Rates (RAMR) presented here adjusts the observed mortality rates. This statistical methodology takes into account pre-existing health problems that put some patients at greater risk of death to level the playing field and allow fair comparisons across hospitals. __`r length(df$Risk.Adjusted.Mortality.Rate[is.na(df$Risk.Adjusted.Mortality.Rate)])`__ missing values.
- Number of Deaths: Number of patients that died in this hospital. __`r length(df$X..of.Deaths[is.na(df$X..of.Deaths)])`__ missing values.
- Number of Cases: Number of patients that had this medical procedure or condition in this hospital. __`r length(df$X..of.Cases[is.na(df$X..of.Cases)])`__ missing values.

__Remove missing values, because number of missing values consists of half of dataset.__
```{r}
df_clean <- df[which(is.na(df$X..of.Cases)==F),]
```

__Clean Dataset:__ __`r nrow(df_clean)`__ observations and __`r ncol(df_clean)`__ variables. 

__Variables with no missing values:__

- Year: two years 2012 and 2013; __`r length(df_clean$Year[df_clean$Year==2012])`__ values for 2012 year and __`r length(df_clean$Year[df_clean$Year==2013])`__ values for 2013 year. 
- County: 55 counties.
- Hospital: 341 hospitals. 
- OSHPDID: A unique number established by the Office of Statewide Health Planning and Development (OSHPD) for identifying facilities and used in the Licensed Facility Information System (LFIS). The first three numbers identify the type of facility, the next two represent the county number, and the last five are randomly assigned within each county. 570261 unique codes. 
- Longitude: Longitude of hospital. 
- Latitude: Latitude of hospital.
- location1: 333 levels. 
- Hospital Ratings: Comparison rating based on a 95% Confidence Interval (CI). If a hospitals upper CI is less than the statewide observed rate, it is designated as performing better than the average hospital. If a hospitals lower CI is greater than the state rate, it is designated as performing worse than the average state hospital. __3 levels of Hospital Ratings:__ As Expected, Better and Worse.
```{r}
summary(df_clean$Hospital.Ratings)
```

- Procedure/Condition: Procedure that was performed or condition that was treated. 6 medical procedures performed: Abdominal Aortic Aneurysm (AAA) Repair, Carotid Endarterectomy, Craniotomy, Esophageal Resection, Pancreatic Resection, Percutaneous Coronary Intervention. 6 medical conditions treated: Acute Stroke, Acute Myocardial Infarction, Heart Failure, Gastrointestinal Hemorrhage, Hip Fracture and Pneumonia. Clean dataset consists of __17 levels, instead of 12.__ 
```{r}
summary(df_clean$Procedure.Condition)
```

__Decoding Procedure/Condition variable.__

According to the American Stroke Association (ASA), strokes can be classified into 2 main categories: __87%__ are ischemic strokes, caused by blockage of an artery; __13%__ are hemorrhagic strokes, caused by bleeding. Ischemic strokes are further divided into 2 groups: thrombotic and embolic strokes. Hemorrhagic strokes are divided into 2 main categories: intracerebral and subarachnoid hemorrhages. 

Our clean dataset has four categories for Acute Stroke: 

- Acute Stroke: 617 observations;
- Acute Stroke Hemorrhagic: 466 observations;
- Acute Stroke Ischemic: 615 obervations;
- Acute Stroke Subarachnoid: 241 observations.

Within each hospital, there are different notations for Acute Stroke variable. It suggests that different doctor uses different notations for the condition. These four categories are combined in one: Acute Stroke. 

```{r}
df_clean$Procedure.Condition <- gsub("Acute Stroke .*","Acute Stroke",df_clean$Procedure.Condition)
df_clean$Procedure.Condition <- factor(df_clean$Procedure.Condition)
```

Two additional categories are present in Procedure/Condition variable:

- Pancreatic Cancer: 142 observations;
- Pancreatic Other: 130 observations. 

These categories are separate medical conditions and are not combined in one category. 

The Procedure.Condition variable contains 6 medical procedures and 8 medical conditions. To indicate what procedure was performed or what condition was treated, the Medical_Categorey variable was added to the clean dataset. 

```{r}
df_clean <- df_clean %>% 
  mutate(Medical_Category = ifelse(grepl("Repair",Procedure.Condition) | grepl("Endarterectomy",Procedure.Condition) | grepl("Craniotomy",Procedure.Condition) | grepl("Resection",Procedure.Condition) | grepl("PCI",Procedure.Condition),               "Procedure", "Condition"))
#glimpse(df_clean)
```

## Explanatory Data Analysis

__Density Plots for # of Cases, # of Deaths and Risk Adjusted Mortality Rate by Hospital Ratings.__ 
```{r,warning=FALSE,message=FALSE}
p1 <- ggplot(df_clean,aes(log(X..of.Cases),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)

p2 <- ggplot(df_clean,aes(log(X..of.Deaths),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)

p3 <- ggplot(df_clean,aes(log(Risk.Adjusted.Mortality.Rate),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)

grid.arrange(p1, p2, p3, ncol=1)
```

__Conclusions 1:__ 

- Distributions between 2012 and 2013 years look simiar (graphs are not shown). 
- There are no associations between variables for number of deaths and number of cases.
- There is association between the risk adjusted mortality rate and hospital ratings.
- Lower the risk adjusted mortality rate, better the hospital ratings.
- Higher the risk adjusted mortality rate, worse the hospital ratings.

__Associations between medical procedures or conditions with hospital ratings, number of cases, number of deaths and risk adjusted mortality rate.__

__Procedures.__
```{r,warning=FALSE,message=FALSE}
df_p <- df_clean[which(df_clean$Medical_Category=="Procedure"),]

df_p_all <- df_p %>% 
  group_by(Procedure.Condition) %>% 
  summarise(all_cases = sum(X..of.Cases), 
            all_deaths = sum(X..of.Deaths),
            all_mortality_rate = sum(Risk.Adjusted.Mortality.Rate))

# df_p_all %>% arrange(desc(all_cases))
# df_p_all %>% arrange(desc(all_deaths))
# df_p_all %>% arrange(desc(all_mortality_rate))

df_p_all 
```

__Conditions.__
```{r,warning=FALSE,message=FALSE}
df_c <- df_clean[which(df_clean$Medical_Category=="Condition"),]

df_c_all <- df_c %>% 
  group_by(Procedure.Condition) %>% 
  summarise(all_cases = sum(X..of.Cases), 
            all_deaths = sum(X..of.Deaths),
            all_mortality_rate = sum(Risk.Adjusted.Mortality.Rate))

#df_c_all %>% arrange(desc(all_cases))
#df_c_all %>% arrange(desc(all_deaths))
#df_c_all %>% arrange(desc(all_mortality_rate))

df_c_all
```

__Hospital Ratings.__
```{r,warning=FALSE,message=FALSE}
prop.table(table(df_clean$Procedure.Condition,df_clean$Hospital.Ratings))*100
```

__Conclusions 2:__

- The highest number of cases is for PCI and Craniotomy procedures, Acute Stroke, Heart Failure, AMI and GI Hemorrhage conditions.
- The highest number of deaths is for Craniotomy and PCI procedures, Acute Stroke, AMI and Heart Failure conditions.
- The highest mortality rates is for Craniotomy and Pancreatic Resection procedures, Acute Stroke, AMI and Heart Failure conditions. 

- The lowest number of cases is for Esophageal Resection procedure, Pancreatic Cancer and Pancreatic Other conditions. 
- The lowest number of deaths is for Esophageal Resection procedure, Pancreatic Cancer and Pancreatic Other conditions.
- The lowest mortality rates is for Carotid Endarterectomy procedure, Pancreatic Other and Pneumonia conditions. 

- Better ratings are for Craniotomy procedure, Acute Stroke, AMI, GI Hemorrhage and Heart Failure conditions.
- Worse ragings are for Craniotomy and PCI procedures, Acute Stroke, AMI, GI Hemorrhage and Heart Failure conditions.  
- As Expected ratings are for Acute Stroke, AMI, GI Hemorrhage and Heart Failure conditions.

__Density Plots for Risk Adjusted Mortality Rate by Procedures Performed and Hospital Ratings.__ 
```{r,warning=FALSE,message=FALSE,eval=FALSE,echo=FALSE}
df_p <- df_clean[which(df_clean$Medical_Category=="Procedure"),]

p4 <- ggplot(df_p,aes(log(X..of.Cases),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p4
```

```{r,warning=FALSE,message=FALSE,eval=FALSE,echo=FALSE}
p5 <- ggplot(df_p,aes(log(X..of.Deaths),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p5
```

```{r,warning=FALSE,message=FALSE}
p6 <- ggplot(df_p,aes(log(Risk.Adjusted.Mortality.Rate),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p6
```

__Density Plots for Risk Adjusted Mortality Rate by Conditions Treated and Hospital Ratings.__
```{r,warning=FALSE,message=FALSE,eval=FALSE,echo=FALSE}
df_c <- df_clean[which(df_clean$Medical_Category=="Condition"),]

p7 <- ggplot(df_c,aes(log(X..of.Cases),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p7
```

```{r,warning=FALSE,message=FALSE,eval=FALSE,echo=FALSE}
p8 <- ggplot(df_c,aes(log(X..of.Deaths),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p8
```

```{r,warning=FALSE,message=FALSE}
p9 <- ggplot(df_c,aes(log(Risk.Adjusted.Mortality.Rate),fill=factor(Hospital.Ratings),colour=factor(Hospital.Ratings)))+
  geom_density(alpha = 0.1)+
  theme(legend.position='bottom')+
  facet_wrap(~ Procedure.Condition, ncol=2, scales="free_y")

p9
```

__Summary:__ The most contributing procedues and conditions. 

- __Procedures__: 
    - the highest: PCI, Craniotomy, Pancreatic Resection.
    - the lowest: Esophageal Resection, Carotid Endarterectomy. 
  
- __Conditions__: 
    - the highest: Acute Stroke, AMI, Heart Failure, GI Hemorrhage.
    - the lowest: Pancreatic Cancer, Pancreatic Other, Pneumonia. 

### __Mapping and summary__ of overall hospital quality ratings and mean mortality rate among all conditions and procedures. 

__Summary of hospital ratings over all conditions and procedues.__ 
```{r,warning=FALSE,message=FALSE}
df_clean <- df_clean %>% mutate(ratings =
           ifelse(grepl("As Expected",Hospital.Ratings),"0",                        
           ifelse(grepl("Better",Hospital.Ratings),"1",
           ifelse(grepl("Worse",Hospital.Ratings),"-1",NA))))
df_clean$ratings <- as.numeric(df_clean$ratings)

all_ratings <- df_clean %>% 
                group_by(Hospital,Latitude,Longitude) %>% 
                summarise(all_ratings = sum(ratings), 
                          mean_mortality_rate = mean(Risk.Adjusted.Mortality.Rate)) %>% 
                mutate(ratings =
                         ifelse(all_ratings > 0,"Better",
                         ifelse(all_ratings < 0, "Worse","As Expected")))
all_ratings$ratings <- as.factor(all_ratings$ratings)
all_ratings <- tbl_df(all_ratings)
```

__Overall Hospital Ratings:__
```{r,warning=FALSE,message=FALSE}
summary(all_ratings$ratings)
```

- Hospitals with __better__ quality ratings:
```{r,warning=FALSE,message=FALSE}
all_ratings %>% arrange(desc(all_ratings)) %>% select(Hospital,all_ratings) %>% slice(1:10)
```

- Hospitals with __worse__ quality ratings:
```{r,warning=FALSE,message=FALSE}
all_ratings %>% arrange(all_ratings) %>% select(Hospital,all_ratings) %>% slice(1:10)
```

- Hospitals with the __highest__ mean mortality rate:
```{r,warning=FALSE,message=FALSE}
all_ratings %>% arrange(desc(mean_mortality_rate)) %>% select(Hospital,mean_mortality_rate)  %>% slice(1:10)
```

- Hospitals with the __lowest__ mean mortality rate:
```{r,warning=FALSE,message=FALSE}
all_ratings %>% arrange(mean_mortality_rate) %>% select(Hospital,mean_mortality_rate) %>% slice(1:10)
```

__Mapping of overall hospital ratings and mean mortality rates.__ 
```{r,warning=FALSE,message=FALSE}
CAmap <- get_map(location="California",source="google",maptype="roadmap",crop=FALSE,zoom=6) 
ggmap (CAmap) +
  geom_point(aes(x=Longitude,y=Latitude,size=mean_mortality_rate,colour=ratings),data=all_ratings,alpha=0.7)+
  scale_colour_manual(values=c("Worse" = "darkred","Better" = "darkblue","As Expected" = "darkgrey"))+
  scale_size(range = c(0, 10))
```

### __Analysis 2:__

I will analyze which condition or procedure has the best or worse hospital ratings, and map these results only for selected conditions or procedures. 

### __Predictions__

To predict hospital ratings I will use classification decision trees, random forests or multinomial logistic regression. First, I will use three models to train them on 2012-2013 data and access which model gives the best performance on the training data, then I will test three model performances on 2014 test data. 

### __Recommendations to Patients__

### Hospitals with the best overall ratings and the lowest mean mortality rate in state of California.
```{r,warning=FALSE,message=FALSE}
best_ratings <- all_ratings %>% arrange(desc(all_ratings)) %>% slice(1:25)
lowest_mort_rate <- all_ratings %>% arrange(mean_mortality_rate) %>% slice(1:25)
best_lowest <- bind_rows(best_ratings,lowest_mort_rate)

CAmap <- get_map(location="California",source="google",maptype="roadmap",crop=FALSE,zoom=6) 
ggmap (CAmap) +
  geom_point(aes(x=Longitude,y=Latitude,size=ratings,colour=mean_mortality_rate),data=best_lowest,alpha=0.7)+
  scale_colour_distiller(type = "div", palette = "RdBu", direction = -1)
```

### Hospitals with ...

2)	Recommend to patients which hospital has the best ratings for particular medical condition or procedure in state of California;

### Hospitals with ...

3)	Recommend which hospital will have the best care in the future using predicted hospital ratings. 





