---
title: "Data Visualization for Titanic Data"
author: "Maria P. Frushicheva"
date: "Sunday, July 31, 2016"
output: html_document
---
Exploratory data analysis for Titanic dataset: investigation whether you'd have a chance of surviving the disaster.

Load the cleaned data into dataframe titanic. 
```{r}
titanic <- read.csv("titanic_clean.csv", header = TRUE, sep = ",")
```

Check out the structure of titanic.
```{r}
str(titanic)
```

The last passanger had missing information in all fields, except for the age variable, therefore was excluded from the dataset. 
```{r}
tail(titanic)
titanic <- titanic[-1310,]
```

Use ggplot() to plot the distribution of sexes within the classes of the ship. 
```{r,warning=FALSE}
require(ggplot2)
ggplot(titanic,aes(x=factor(pclass),fill=factor(sex)))+
  geom_bar(position="dodge")
```

Use ggplot() to estimate your chances of survival from the distribution of sexes within the classes of the ship. 
```{r}
ggplot(titanic,aes(x=factor(pclass),fill=factor(sex)))+
  geom_bar(position="dodge")+
  facet_grid(". ~ survived")
```

Use ggplot() to estimate your chances of survival based on your age from the distribution of sexes within the classes of the ship. 
```{r}
posn.j <- position_jitter(0.5, 0)
ggplot(titanic,aes(x=factor(pclass),y=age,col=factor(sex)))+
  geom_jitter(size=3,alpha=0.5,position=posn.j)+
  facet_grid(". ~ survived")
```