---
title: "Titanic Exercise"
author: "Maria P. Frushicheva"
date: "Sunday, July 31, 2016"
output: html_document
---
Dealing with missing values in Titanic dataset. 

Load the titanic3.xls data into dataframe titanic. 
```{r,warning=FALSE}
require(xlsx)
lapply("titanic3.xls", function(f) {
  df = read.xlsx(f, sheetName = "titanic3")
  write.csv(df, gsub(".xls","_original.csv", f), row.names=FALSE)
})
titanic <- read.csv("titanic3_original.csv", header = TRUE, sep = ",")
```

Installing packages for tidy data.
```{r,warning=FALSE}
# install.packages("dplyr")
# install.packages("tidyr")
require(dplyr)
require(tidyr)
df <- tbl_df(titanic)
```

Port of embarkation variable: find the missing values and replace them with S.
```{r}
df$embarked[is.na(df$embarked)] <- "S"
```

Age variable: use mean (or median) value to populate the missing values.
```{r}
df %>% summarise(Min = min(age, na.rm=TRUE),
                     Median = median(age, na.rm=TRUE),
                     Mean = mean(age, na.rm=TRUE),
                     Max = max(age, na.rm=TRUE))

age_mean <- mean(df$age, na.rm=TRUE)
df$age[is.na(df$age)] <- age_mean
```

Boat variable: fill missing values with a dummy value, e.g. the string 'None' or 'NA'.
```{r}
class(df$boat)
df$boat <- as.character(df$boat)
df$boat[is.na(df$boat)] <- "NA"
```

Cabin variable: the cabin number is missing might be a useful indicator of survival, thus create a new column has_cabin_number which has 1 if there is a cabin number, and 0 otherwise. 
```{r}
df <- df %>% mutate(has_cabin_number = ifelse(!is.na(cabin),1,0))
```
Write the data in cleaned up csv file. 
```{r}
write.csv(df, "titanic_clean.csv", row.names=FALSE)
```



